Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-08-01 paper.tex
minor cgange
updated figures and text, new masses

%\voffset-.6in 
\voffset.3in

@@

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}
\newcommand{\sm}{\rmn{sm}}
\newcommand{\mpc}{\rmn{Mpc}}
\newcommand{\kpc}{\rmn{kpc}}
\newcommand{\mev}{\rmn{MeV}}
\newcommand{\cm}{\rmn{cm}}
\newcommand{\egt}{\tilde{E}_\gamma}
\newcommand{\eet}{\tilde{E}_\rmn{e}}
\newcommand{\bra}{\langle}
\newcommand{\ket}{\rangle}
%OLD
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\gev}{\rmn{GeV}}
\newcommand{\tev}{\rmn{TeV}}
\newcommand{\ev}{\rmn{eV}}
\newcommand{\dd}{\rmn{d}}
\newcommand{\mx}{\ensuremath{m_{\chi}}}
\newcommand{\ga}{\gamma}
\newcommand{\ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\ngammaj}{\ensuremath{N_{\gamma,j}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\sigv}{v_\rmn{cl}}
\newcommand{\ic}{\rmn{IC}}
\newcommand{\egamma}{\ensuremath{E_{\gamma}}}
\newcommand{\dic}{\rmn{dIC}}
\newcommand{\fsr}{\rmn{FSR}}
\newcommand{\CR}{\rmn{CR}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\rvir}{r_{200}}
\newcommand{\mvir}{M_{200}}
\newcommand{\rhoc}{\ensuremath{\rho_c}}
\newcommand{\e}{\rmn{e}}
\newcommand{\eg}{E_\gamma}
\newcommand{\ir}{\rmn{ir}}

\voffset.8in

@@

\begin{abstract}
  We study the possibility for detecting gamma-ray emission in galaxy
  clusters. We consider 1) cosmic ray (CR) induced pion decay which is
  thought to dominate the astrophysical signal from clusters, 2)
  different representative benchmark models of supersymmetric dark
  matter (DM), and 3) leptophilic models of DM annihilation that
  recently have been proposed to explain the positron excess seen by
  PAMELA and include a Sommerfeld enhancement of the particle physics
  cross section over its standard value. To model DM annihilation, we
  consider continuum emission as a result of the hadronization of
  annihilating neutralinos, internal bremsstrahlung, and inverse
  Compton emission from the cosmic microwave background as well as
  from a realistic spatial and spectral distribution of dust and
  stellar light. We predict the Virgo and Fornax clusters to be the
  brightest DM sources and find a particularly low CR induced
  background for Fornax. For a minimum substructure mass given by the
  DM free-streaming scale, the largest DM halo mass maximizes the
  substructure boost over the smoothly distributed annihilation flux:
  for galaxy clusters as the largest collapsed objects, we find a
  boost factor of more than 1000. Since the annihilation flux of
  substructures is mostly contributed by the regions around the virial
  radius, the resulting surface brightness profiles are almost
  flat. This makes it very challenging to detect this flux with
  imaging air Cherenkov telescopes since their sensitivity drops
  linearly with radius and they typically have $5-10$ linear
  resolution elements across a cluster; the gamma-ray satellite \Fermi
  is better suited. Assuming cold dark matter with a substructure mass
  distribution down to an Earth mass and using \Fermi upper limits, we
  rule out the leptophilic models in their present form in 26
  clusters, and limit the boost from Sommerfeld enhancement in the
  most constraining cluster M49 to be less than 14. This corresponds
  to a Sommerfeld enhancement in the Milky Way of less than
  8. Alternatively, if the Sommerfeld enhancement factor is realized
  in Nature, this would imply a limiting substructure mass in M49 of
  $\mlim > 10\,\msun$ which would be a challenge for structure
  formation. \C{We find it challenging for \Fermi to
    constrain our selection of DM benchmark models using clusters
    without improved modelling of extended sources in addition to a
    stacked likelihood analysis.} The \Fermi upper limits are, however,
  closing in on our predictions for the CR flux using an analytic
  model based on cosmological hydrodynamical cluster simulations. Thus
  we will soon start to constrain the underlying CR physics such as
  shock acceleration efficiencies or CR transport properties.
\end{abstract}

@@

3. Galaxy clusters. This possibility has been less studied \C{theoretically}, however
\C{currently there is an ongoing observational campaign to detect
  gamma-ray emission from galaxy clusters
  \cite{2010ApJ...710..634A,2010JCAP...05..025A,2009A&A...495...27A,2009IJMPD..18.1627D,2009A&A...502..437A,2009arXiv0907.5000G,2009ApJ...704..240K,2009ApJ...706L.275A,2010ApJ...717L..71A}. In fact, we noted} in a previous Letter
\cite{2009PhRvL.103r1302P} that there are certain advantages that work
in favour of this possible target for gamma-ray detection of DM
annihilation. Galaxy clusters constitute the most massive objects in
our Universe that are forming today. This causes their DM subhalo mass
function to be less affected by tidal stripping compared to galaxy
sized halos that formed long ago.  The annihilation luminosities of
the DM halo component for e.g. the Virgo cluster and the Draco dwarf
scales in a way (see \cite{2009PhRvL.103r1302P}) that the ratio of
gamma-ray luminosities from the smooth components is around 4, in
favour of Virgo. In addition, there may be a further enhancement due
to substructure, which to a large extent should be unaffected by tidal
disruption, at least in the outer regions.  According to a recent
estimate \cite{2011MNRAS.410.2309G}, more massive haloes tend to have
a larger mass fraction in subhalos.  For example, cluster size haloes
typically have 7.5 per cent of the mass within $r_{200}$ in
substructures of fractional mass larger than $10^{-5}$, which is 25
per cent higher than for galactic haloes.\footnote{All halo masses and
  length scales in this paper are scaled to the currently favored
  value of Hubble's constant, $H_0 = 70\,
  \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define the virial mass $\mvir$
  and virial radius $\rvir$ as the mass and radius of a sphere
  enclosing a mean density that is 200 times the critical density of
  the Universe $\rho_{\rmn{cr}}$.}

@@

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize already
in Table~\ref{tab:flux_tab} our results for the gamma-ray flux from
four promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. If the boost from Sommerfeld enhancement (SFE)
is small, however, then the gamma-ray emission induced by CRs is
dominating most clusters with the exception of a few clusters where
Fornax had the highest flux from DM. In particular, we find Fornax and
Virgo as the prime targets for DM observations, Perseus for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma for comparison. While studying these four clusters in
great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy
clusters from the Rosat all-sky survey (extended HIghest X-ray FLUx
Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R}).\footnote{Note that we have added the
  Virgo cluster to the sample that we nevertheless refer to as
  extended HIFLUGCS catalogue in the following.}  \C{For previous work
  related to DM in clusters, see, e.g.,
  \cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
  and for recent work on CRs in clusters see
  \cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.}

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters within $\rvir$.}
\begin{tabular}{l  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ $^{(1)}$ & $\B_\rmn{sfe} $$^{(2)}$ &  $\B_\rmn{sub} $$^{(3)}$ \\
         & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ 
         & & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Coma                 & $5.1\times10^{-8}$  & $8.2\times10^{-11}$ & $4.0\times10^{-9}$  
 & \,\,\,\,\,         & $7.4\times10^{-12}$ & $3.0\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $13.8$ & $530/60$  & $1290$ \\
 Perseus \,\,\,\,\,\, & $4.4\times10^{-8}$  & $7.1\times10^{-11}$ & $1.4\times10^{-8}$  
 & \,\,\,\,\,         & $6.4\times10^{-12}$ & $2.6\times10^{-13}$ & $5.2\times10^{-12}$ 
 & $7.71$ & $530/70$  & $1210$ \\
 Virgo                & $9.2\times10^{-7}$  & $1.4\times10^{-9}$ & $1.1\times10^{-8}$  
 & \,\,\,\,\,         & $1.3\times10^{-10}$ & $5.4\times10^{-12}$ & $3.9\times10^{-12}$ 
 & $6.9$  & $530/105$  & $1110$  \\
 Fornax               & $7.9\times10^{-8}$  & $1.3\times10^{-10}$ & $2.2\times10^{-10}$ 
 & \,\,\,\,\,         & $1.1\times10^{-11}$ & $4.6\times10^{-13}$ & $8.3\times10^{-14}$ 
 & $1.0$  & $530/130$  & $730$  \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) \C{The mass of Fornax, Coma, and Perseus are taken from \cite{2002ApJ...567..716R}, 
  while the mass of Virgo is derived from \cite{1984ApJ...281...31T}.}\\
  (2) \C{The boost due to Sommerfeld enhancement. The first value shows the saturated 
  Sommerfeld boost realized when substructures are present, the latter is the Sommerfeld 
  boost without substructures. See Section II A 1.}\\
  (3) \C{We integrate the emission in a cylinder with a radial extent of $2.5
  R_{200}$ along the line-of-sight and an angular size corresponding to $\rvir$
  (both measured from the cluster center).}\\
  (4) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  \Fermi (see the following sections for detail).\\
  (5) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (6) Gamma-ray flux induced by CR protons. See Section V C. 
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and \Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) caused intense
activity in the theoretical DM community. At first, there was
scepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld enhancement
(SFE). This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
\begin{equation}
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
\end{equation}
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
\begin{equation}
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
\label{eq:saturation}
\end{equation}
In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$\sigv$ the SFE will be smaller than the one expected in a single
galaxy such as the MW, roughly by a factor $v_{\rm MW}/\sigv$. \C{Note
  that the $1/v$-scaling is valid only for $(v/c) \gtrsim
  (m_\phi/m_\chi)$; at smaller velocities and outside resonances, the
  $1/v$-enhancement saturates at $m_\phi/m_\chi$
  \cite{2008PhRvL.101z1301K}.}

@@

processes; leptophilic DM annihilating \C{through

@@

  \mbox{~or~}4\pi\to4e\}$ (neglecting the produced neutrinos) where
  the $\rmn{e}^+/\rmn{e}^-$ pairs} IC upscatter background photons
(LP-IC), leptophilic DM emitting final state radiation (LP-FSR),
supersymmetric DM BM models where annihilating neutralinos generate

@@

1+\rho_{\sub,i}^2(r)/\rho^2(r)$ \C{\footnote{Note that if the boost

@@

    approaches the constant saturated boost of 530.}} (see

@@

%-) Consider having one figure where we compare the Einasto profile, to
%NFW, NFW w/ beta 0.6, and also the effect of a different concentration
%mass-scaling relation
%
%-) also, plot BM flux vs. psf, for models w/ and w/o substructures

@@

PERSEUS  &  68 &  56.14 &  75.31 &  20.64 &  63.95 &  13.92 &  24.52 \\
OPHIUCHUS &  96 &  23.31 &  31.27 &   8.57 &  26.55 &   5.88 &  10.28 \\
COMA     &  34 &  17.50 &  23.48 &   6.43 &  19.94 &   2.62 &   7.56 \\
NORMA    &  94 &  13.39 &  17.97 &   4.92 &  15.26 &   1.16 &   5.57 \\
CENTAURUS &  30 &  12.06 &  16.17 &   4.43 &  13.73 &   2.45 &   5.04 \\

>>>>

%\voffset-.6in 
%\voffset.3in

@@

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}
\newcommand{\sm}{\rmn{sm}}
\newcommand{\mpc}{\rmn{Mpc}}
\newcommand{\kpc}{\rmn{kpc}}
\newcommand{\mev}{\rmn{MeV}}
\newcommand{\cm}{\rmn{cm}}
\newcommand{\egt}{\tilde{E}_\gamma}
\newcommand{\eet}{\tilde{E}_\rmn{e}}
\newcommand{\bra}{\langle}
\newcommand{\ket}{\rangle}
\newcommand{\vst}{\vspace{-0.14mm}}
\newcommand{\el}{\rmn{el}}

%OLD
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\gev}{\rmn{GeV}}
\newcommand{\tev}{\rmn{TeV}}
\newcommand{\ev}{\rmn{eV}}
\newcommand{\dd}{\rmn{d}}
\newcommand{\mx}{\ensuremath{m_{\chi}}}
\newcommand{\ga}{\gamma}
\newcommand{\ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\ngammaj}{\ensuremath{N_{\gamma,j}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\sigv}{v_\rmn{cl}}
\newcommand{\ic}{\rmn{IC}}
\newcommand{\egamma}{\ensuremath{E_{\gamma}}}
\newcommand{\dic}{\rmn{dIC}}
\newcommand{\fsr}{\rmn{FSR}}
\newcommand{\CR}{\rmn{CR}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\rvir}{r_{200}}
\newcommand{\mvir}{M_{200}}
\newcommand{\rhoc}{\ensuremath{\rho_c}}
\newcommand{\e}{\rmn{e}}
\newcommand{\eg}{E_\gamma}
\newcommand{\ir}{\rmn{ir}}

%\voffset.8in
\voffset.4in

@@

\begin{abstract}
  We study the possibility for detecting gamma-ray emission in galaxy
  clusters. We consider 1) cosmic ray (CR) induced pion decay which is
  thought to dominate the astrophysical signal from clusters, 2)
  different representative benchmark models of supersymmetric dark
  matter (DM), and 3) leptophilic models of DM annihilation that
  recently have been proposed to explain the positron excess seen by
  PAMELA and include a Sommerfeld enhancement of the particle physics
  cross section over its standard value. To model DM annihilation, we
  consider continuum emission as a result of the hadronization of
  annihilating neutralinos, internal bremsstrahlung, and inverse
  Compton emission from the cosmic microwave background as well as
  from a realistic spatial and spectral distribution of dust and
  stellar light. We predict the Virgo, Fornax and \C{M49} clusters to
  be the brightest DM sources and find a particularly low CR induced
  background for Fornax. For a minimum substructure mass given by the
  DM free-streaming scale, the largest DM halo mass maximizes the
  substructure boost over the smoothly distributed annihilation flux:
  for galaxy clusters as the largest collapsed objects, we find a
  boost factor of more than 1000. Since the annihilation flux of
  substructures is mostly contributed by the regions around the virial
  radius, the resulting surface brightness profiles are almost
  flat. This makes it very challenging to detect this flux with
  imaging air Cherenkov telescopes since their sensitivity drops
  linearly with radius and they typically have $5-10$ linear
  resolution elements across a cluster; the gamma-ray satellite \Fermi
  is better suited. Assuming cold dark matter with a substructure mass
  distribution down to an Earth mass and using \Fermi upper limits, we
  rule out the leptophilic models in their present form in 28
  clusters, and limit the boost from Sommerfeld enhancement in the
  most constraining clusters, \C{M49 and Fornax, } to be less than 5. This
  corresponds to a Sommerfeld enhancement in the Milky Way of less
  than 5. Alternatively, if the Sommerfeld enhancement factor is
  realized in Nature, this would imply a limiting substructure mass of
  $\mlim > 10^4\,\msun$ which would be a challenge for structure
  formation. We find it challenging for \Fermi to constrain our
  selection of DM benchmark models in the \C{near future} using clusters
  without improved modelling of extended sources in addition to a
  stacked likelihood analysis. The \Fermi upper limits are, however,
  closing in on our predictions for the CR flux using an analytic
  model based on cosmological hydrodynamical cluster simulations. Thus
  we will soon start to constrain the underlying CR physics such as
  shock acceleration efficiencies or CR transport properties.
\end{abstract}

@@

3. Galaxy clusters. This possibility has been less studied
theoretically, however currently there is an ongoing observational
campaign to detect gamma-ray emission from galaxy clusters
\cite{2010ApJ...710..634A,2010JCAP...05..025A,2009A&A...495...27A,2009IJMPD..18.1627D,2009A&A...502..437A,2009arXiv0907.5000G,2009ApJ...704..240K,2009ApJ...706L.275A,2010ApJ...717L..71A}. In
fact, we noted in a previous Letter \cite{2009PhRvL.103r1302P} that
there are certain advantages that work in favour of this possible
target for gamma-ray detection of DM annihilation. Galaxy clusters
constitute the most massive objects in our Universe that are forming
today. This causes their DM subhalo mass function to be less affected
by tidal stripping compared to galaxy sized halos that formed long
ago.  The annihilation luminosities of the DM halo component for
e.g. the Virgo cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions.  According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos.  For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than for galactic haloes.\footnote{All halo masses and length scales
  in this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define
  the virial mass $\mvir$ and virial radius $\rvir$ as the mass and
  radius of a sphere enclosing a mean density that is 200 times the
  critical density of the Universe $\rho_{\rmn{cr}}$.}

@@

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize already
in Table~\ref{tab:flux_tab} our results for the gamma-ray flux from
four promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. If the boost from Sommerfeld enhancement (SFE)
is small, however, then the gamma-ray emission induced by CRs is
dominating most clusters with the exception of a few clusters where
Fornax had the highest flux from DM. In particular, we find Fornax and
Virgo as the prime targets for DM observations, Perseus for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma for comparison. While studying these four clusters in
great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy
clusters from the Rosat all-sky survey (extended HIghest X-ray FLUx
Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
  we have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.}  For
previous work related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
and for recent work on CRs in clusters see
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters within $\rvir$.}
\begin{tabular}{l  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ $^{(1)}$ & $\B_\rmn{sfe} $$^{(2)}$ &  $\B_\rmn{sub} $$^{(3)}$ \\
         & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ 
         & & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Coma                 & $8.2\times10^{-8}$  & $7.6\times10^{-11}$ & $4.1\times10^{-9}$  
 & \,\,\,\,\,         & $1.2\times10^{-11}$ & $2.8\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $12.9$ & $530/65$  & $1290$ \\
 Perseus \,\,\,\,\,\, & $8.6\times10^{-8}$  & $7.8\times10^{-11}$ & $1.5\times10^{-8}$  
 & \,\,\,\,\,         & $1.2\times10^{-11}$ & $2.9\times10^{-13}$ & $5.5\times10^{-12}$ 
 & $8.6$ & $530/75$  & $1190$ \\
 Virgo                & $1.6\times10^{-6}$  & $1.5\times10^{-9}$ & $1.5\times10^{-8}$  
 & \,\,\,\,\,         & $2.3\times10^{-10}$ & $5.4\times10^{-12}$ & $5.7\times10^{-12}$ 
 & $6.9$  & $530/80$  & $1120$  \\
 Fornax               & $3.5\times10^{-7}$  & $3.2\times10^{-10}$ & $3.1\times10^{-10}$ 
 & \,\,\,\,\,         & $5.1\times10^{-11}$ & $1.2\times10^{-12}$ & $1.1\times10^{-13}$ 
 & $2.4$  & $530/110$  & $890$  \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The mass of Fornax, Coma, and Perseus are derived from \cite{2007A&A...466..805C}, 
  while the mass of Virgo is derived from \cite{1984ApJ...281...31T}.\\
  (2) The boost due to Sommerfeld enhancement. The first value shows the saturated 
  Sommerfeld boost realized when substructures are present, the latter is the Sommerfeld 
  boost without substructures. See Section II A 1.\\
  (3) We integrate the emission in a cylinder with a radial extent of $2.5
  R_{200}$ along the line-of-sight and an angular size corresponding to $\rvir$
  (both measured from the cluster center).\\
  (4) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  \Fermi (see the following sections for detail).\\
  (5) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (6) Gamma-ray flux induced by CR protons. See Section V C. 
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and \Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) caused intense
activity in the theoretical DM community. At first, there was
scepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld enhancement
(SFE). This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
\begin{equation}
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
\end{equation}
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
\begin{equation}
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
\label{eq:saturation}
\end{equation}
In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$\sigv$ the SFE will be smaller than the one expected in a single
galaxy such as the MW, roughly by a factor $v_{\rm MW}/\sigv$. Note
that the $1/v$-scaling is valid only for $(v/c) \gtrsim
(m_\phi/m_\chi)$; at smaller velocities and outside resonances, the
$1/v$-enhancement saturates at $m_\phi/m_\chi$
\cite{2008PhRvL.101z1301K}.

@@

processes; leptophilic DM annihilating through

@@

\mbox{~or~}4\pi\to4e\}$ (neglecting the produced neutrinos) where the
$\rmn{e}^+/\rmn{e}^-$ pairs IC upscatter background photons (LP-IC),
leptophilic DM emitting final state radiation (LP-FSR), supersymmetric
DM BM models where annihilating neutralinos generate

@@

1+\rho_{\sub,i}^2(r)/\rho^2(r)$ \footnote{Note that if the boost

@@

    approaches the constant saturated boost of 530.} (see

@@

@@

PERSEUS  &  68 &  14.28 &  19.15 &  52.49 &  16.26 &  16.25 &  10.06 \\
OPHIUCHU &  96 &   7.71 &  10.34 &  28.33 &   8.78 &   8.78 &   6.01 \\
COMA     &  34 &   4.15 &   5.56 &  15.24 &   4.72 &   4.72 &   2.55 \\
NORMA    &  94 &   2.98 &   3.99 &  10.94 &   3.39 &   3.38 &   2.19 \\
CENTAURU &  30 &   2.61 &   3.50 &   9.58 &   2.97 &   2.97 &   2.62 \\
TRIANGUL &  95 &   1.77 &   2.37 &   6.50 &   2.01 &   2.01 &   0.37 \\
